---
title: "Biostatistics Final Exam"
author: "Abdalla Mohamed (Matriculation no. 834120)"
date: "2026-02-24"
output: html_document
---

2) 
a. Download the SPSS data file KiGGS03_06.sav from Moodle and import it into R. [2 points] 

```{r}
# Reading the data using foreign library
df <- foreign::read.spss("Raw_Data/KiGGS03_06.sav", to.data.frame = TRUE)
```

b. Create a new dataframe in R named kiggs_xyz (where you should replace xyz with your name) which contains the following variables (and only these): E070M (mother smoking), E070V (father smoking), E072 (mother smoking in pregnancy), E074 (mother smoking while breastfeeding), sex (sex), age2 (age), schichtz (social class), e065z (total sleep per day), and HbA1c (glycohemoglobin; long-term blood glucose marker). [4 points] 

```{r}
# Creation of a new dataframe in R named kiggs_xyz
kiggs_Abdalla <- df[, c("E070M", "E070V", "E072", "E074", "sex", "age2", "schichtz" , "e065z" , "HbA1c")]

```

c. Run the formatting steps in the provided Rmd file data_formatting.Rmd, adapted to your dataframe. Save this formatted dataframe as an RData file on your computer, e.g. on your desktop. [1 point] 

```{r}
# Run the formatting code
kiggs_Abdalla$E070M    <- factor(kiggs_Abdalla$E070M, labels = c("yes, daily", "yes, occasionally", "no"))
kiggs_Abdalla$E070V    <- factor(kiggs_Abdalla$E070V, labels = c("yes, daily", "yes, occasionally", "no"))
kiggs_Abdalla$E072     <- factor(kiggs_Abdalla$E072, labels = c("yes, regularly", "yes, from time to time", "no, never"))
kiggs_Abdalla$E074     <- factor(kiggs_Abdalla$E074, labels = c("yes, regularly", "yes, from time to time", "no, never", "has not breastfed"))
kiggs_Abdalla$sex      <- factor(kiggs_Abdalla$sex, labels = c("boys", "girls"))
kiggs_Abdalla$age2     <- factor(kiggs_Abdalla$age2, labels = c("0-1y", "2-3y", "4-5y", "6-7y", "8-9y", "10-11y", "12-13y", "14-15y", "16-17y"))
kiggs_Abdalla$schichtz <- factor(kiggs_Abdalla$schichtz, labels = c("low social status", "medium social status", "high social status"))

# Saving the formatted file on my desktop clone of the github project
save(kiggs_Abdalla, file = "Formatted_Data/kiggs_Abdalla.RData")

```

3) Question 3 - Data transformations and data checks [13 points] 

a. In order to avoid measuring smoking with 4 variables, we will combine them in one variable called burdenS. This new variable shall contain the total smoke exposure to which the children were exposed to. Carry out the following steps:

o Check that the variables E070M, E070V, E072, E074 are all factors. If they are not, transform them into factors. [2 points] 
o Set the value "has not breastfed" of variable E074 to NA for all children. [1 point] 
o Delete this now empty factor level from the variable. [1 point] 
o Check whether these two steps worked as intended. [1 point] 
o Now calculate the new variable burdenS as the sum of the ranks of the four variables E070M, E070V, E072, E074 for each person (i.e. sum of the numerical factor levels). [4 points] 

What is the meaning of this new variable, does a high value mean that the children were exposed to high levels of smoking, or that they were exposed to low levels of smoking? [1 point] 

```{r}
# o Check that the variables E070M, E070V, E072, E074 are all factors. If they are not, transform them into factors.
sapply(kiggs_Abdalla[,c("E070M", "E070V", "E072", "E074")], is.factor)

# o Set the value "has not breastfed" of variable E074 to NA for all children.
kiggs_Abdalla$E074[which(kiggs_Abdalla$E074 == "has not breastfed")] <- NA

# o Delete this now empty factor level from the variable. [1 point] 
kiggs_Abdalla$E074 <- droplevels(kiggs_Abdalla$E074)

# o Check whether these two steps worked as intended. [1 point] 
table(kiggs_Abdalla$E074)
is.factor(kiggs_Abdalla$E074)

# o Now calculate the new variable burdenS as the sum of the ranks of the four variables E070M, E070V, E072, E074 for each person (i.e. sum of the numerical factor levels). [4 points] 
burdenS <- rowSums(sapply(kiggs_Abdalla[, c("E070M", "E070V", "E072", "E074")], as.numeric))

# What is the meaning of this new variable, does a high value mean that the children were exposed to high levels of smoking, or that they were exposed to low levels of smoking? [1 point]

# Answer: It means that they were exposed to low levels of smoking, as "yes" values ranked first (lower), and "no" values ranked last (higher). So, "12" as example, means it ranked 3 in all the four columns, which means that the children were exposed to the lowest levels of smoking and "4" means the opposite. But, the problem is with "NA" values as they mask some valuable inputs, as one column with NA value, mask the whole column. So, if we need these missed values, we should try different approach of calculating burdenS, but I won't do this in this step as there is no instructions to do so.
```

3) b. Add this variable burdenS to your dataset kiggs_xyz, and save it in its updated form as an RData file (overwrite the previous file). [2 points] 

```{r}
kiggs_Abdalla$burdenS <- rowSums(sapply(kiggs_Abdalla[, c("E070M", "E070V", "E072", "E074")], as.numeric))

save(kiggs_Abdalla, file = "Formatted_Data/kiggs_Abdalla.RData")
```

3) c. What could be an alternative way to summarize the four variables into a summary variable? [1 point]

```{r}
# To use burdenS, I need to deal with NAs, so I will check first no. of rows that have two NAs or more
sum(is.na(kiggs_Abdalla$burdenS))

table(rowSums(is.na(kiggs_Abdalla[, c("E070M", "E070V", "E072", "E074")])))
# 0 in this table means the rows of no missing values, so there are 4187 rows with only one missing value.

# Using dummies variables, by creating 4 columns represents each variable, 3 for the actual values and one for NAs. This will solve the problem of the masking NA. After that, converting these binary codes to decimal could be reflecting meaningfully to the data but in different way. Not from higher and lower, but ranges like "1000,0000,0000,0000" = (This means that the first variable "E070M" has the first value "yes" and the remaining 3 variables are NA values).

# Furthermore, if this binary number converted to decimal number, this means that any number more than 32768 means the first variable "E070M" has "yes" in this column. Also, the maximum exposure means 2^15 + 2^11 + 2^7 + 2^3 = 34952. So, the range between 32768 to 34952 represents the range between maximum exposure in the first variable only (32768) to maximum exposure in all variables (34952). If we have 33160, this could be converted to this binary code (1000,0001,1000,1000) (the "," is imaginary just for explaining), and this binary code means:
# "E070M" = "yes"
# "E070V" = NA
# "E072" = "yes"
# "E074" = "yes"

# This approach gives more information than the first approach and keep full information about the variables values and summarize them.

# -------------------------------------------------------------------------------------
# But there is another approach without Dummies and more relevant to the approach used by numeric and could be useful to use burdenS later as a variable in linear regression
kiggs_Abdalla_n <- kiggs_Abdalla

# As we have 4991 NAs in burdenS, let's check the NA sum per our 4 columns
kiggs_Abdalla_n$fourCol_NA <- rowSums(is.na(kiggs_Abdalla_n[, c("E070M", "E070V", "E072", "E074")]))

table(kiggs_Abdalla_n$fourCol_NA)

# Now, we can remove the rows with 2, 3, and 4 NAs and replace the one NA column with 0
# This will lead to 804 deleted rows, and 4187 new columns with unmasked burdenS values
kiggs_Abdalla_n <- kiggs_Abdalla_n[kiggs_Abdalla_n$fourCol_NA <= 1, ]

kiggs_Abdalla_n[c("E070M", "E070V", "E072", "E074")] <- lapply(kiggs_Abdalla_n[c("E070M", "E070V", "E072", "E074")], function(x) { x <- addNA(x); levels(x)[is.na(levels(x))] <- "0"; x })

# checking the levels of 0 in the four columns to ensure the recalculation
lapply(kiggs_Abdalla_n[, c("E070M", "E070V", "E072", "E074")], levels)

# Now, recalculate burdenS
kiggs_Abdalla_n$burdenS <- rowSums(sapply(kiggs_Abdalla_n[, c("E070M", "E070V", "E072", "E074")], as.numeric))

# We don't need the fourCol_NA, so I will remove it
kiggs_Abdalla_n <- kiggs_Abdalla_n[-(ncol(kiggs_Abdalla_n))]

# First checking the classes of the variables
lapply(kiggs_Abdalla_n[,c("HbA1c","burdenS", "age2", "schichtz", "age2")], class)

kiggs_Abdalla_n <- as.numeric(as.character(kiggs_Abdalla_n[,"HbA1c"]))
```

4) Question 4 – Linear and logistic regression [25 points]

The aim is now to test whether selected variables are associated with HbA1c levels. You may use the previously extracted or derived variable burdenS, but you may also choose other variables.

a) Think about 3 variables from the KiGGS data set that you think may be related to HbA1c. Which variables did you choose? [1 point]

```{r}
# The 3 variables are: burdenS, age2, schichtz (they aren't dependent, but I think checking all the other assumptions required for linear regression isn't the main goal of this question. So, I will just choose the required 3 variables for now.) 

# Test Correlation here #################

# RERWITE THIS AFTER CHOOSING THE VARIABLES
# After checking the continuity of the Y (HbA1c), and the (NA)s, I am choosing "age2", "bmiB" and "OW" variables as they match the assumptions of linear regression. This means that they are independent, no multicollinearity, the NA values can be ignored (147, less than 0.8% of the dataset), so no need for deep imputation.
```



b) 
1- Calculate one linear regression model with all variables as predictors and test in this model which of these 3 variables have an association with HbA1c. [3 points] 

```{r}
# Before linear regression calculation, Let us check the pattern of missing values
mice::md.pattern(kiggs_Abdalla, rotate.names=TRUE)
VIM::aggr(kiggs_Abdalla, prop = FALSE, numbers = TRUE)

# So, I will create two different datasets with two models, one imputated with mice
# and the other is a complete cases, and compare them with my approach up there
# Convert this factor to numeric first in both dataframes
kiggs_Abdalla$HbA1c <- as.numeric(as.character(kiggs_Abdalla$HbA1c))
# df$HbA1c <- as.numeric(as.character(df$HbA1c))
kiggs_Abdalla_cc <- na.omit(kiggs_Abdalla[, c("burdenS", "age2", "schichtz", "HbA1c")])

#kiggs_Abdalla_imp <- df[, c("burdenS", "age2", "schichtz", "HbA1c")]
# Run the imputation on the kiggs dataset
# Extract only the columns you want from the long format
kiggs_Abdalla_imp <- mice::mice(kiggs_Abdalla, m = 5, maxit = 5, meth = 'pmm', 
                                seed = 1, printFlag = TRUE)

mice::densityplot(kiggs_Abdalla_imp)

kiggs_Ab_imp_com <-mice::complete(kiggs_Abdalla_imp)

# Checking imputation
table(!is.na(kiggs_Ab_imp_com$HbA1c), !is.na(kiggs_Ab_imp_com$burdenS), dnn = c("HbA1c","burdenS"))

table(!is.na(kiggs_Ab_imp_com$HbA1c), !is.na(kiggs_Ab_imp_com$age2), dnn = c("HbA1c", "age2"))

table(!is.na(kiggs_Ab_imp_com$HbA1c), !is.na(kiggs_Ab_imp_com$schichtz), dnn = c("HbA1c", "schichtz"))

# First checking the classes of the variables
lapply(kiggs_Ab_imp_com[,c("HbA1c","burdenS", "age2", "schichtz", "age2")], class)

# Now, recalculate burdenS
kiggs_Ab_imp_com$burdenS <- rowSums(sapply(kiggs_Ab_imp_com[, c("E070M", "E070V", "E072", "E074")], as.numeric))

#Running the linear regression model on the 4 created datasets:
#-------------------------------------------------------------
# Running the linear regression model and summerizing it
fit <- lm(HbA1c ~ burdenS + age2 + schichtz, data=kiggs_Abdalla)
summary(fit)

fit_n <- lm(HbA1c ~ burdenS + age2 + schichtz, data=kiggs_Abdalla_n)
summary(fit_n)

fit_cc <- lm(HbA1c ~ burdenS + age2 + schichtz, data=kiggs_Abdalla_cc)
summary(fit_cc)

fit_imp <- lm(HbA1c ~ burdenS + age2 + schichtz, data=kiggs_Ab_imp_com)
summary(fit_imp)


# ----------------------------------------------------------
# Conclusion:
# While all my R^2 values are low < 0.006, suggesting HbA1c is influenced by many other factors that aren't in my models, the Adjusted R^2 is highest in the mice imputed model.

# kiggs_Abdalla (Complete Cases): 0.0073
# kiggs_Ab_imp_com (Imputed): 0.0115

# So, by including the missing data points, the model is explaining about 57% more of the variance in HbA1c than the original model.

# By looking at the Standard Error column for a variable like schichtz (high social status):

# Complete Cases: 0.0123
# Imputed Model: 0.0092

# Smaller standard errors mean my estimates are more precise. Because the imputed model uses 17,628 degrees of freedom instead of 10,008, the "noise" around my estimates decreases.

# The F-statistic tests if any of the used predictors are useful.
# Complete Cases: 7.749
# Imputed Model: 19.71

# This shows a much larger F-statistic (and a significantly smaller p-value: < 2.2e-16) indicates that the relationship between my predictors and HbA1c is much more robust.

# I also tried "as.numeric(age2) + as.numeric(schichtz)" to see the groups total summary and there were no big differences in the conclusion. (Schichtz and age were statistically significant and burdenS wasn't)

# The final conclusion is: I will use the imputated datased 
```

2- Report the p-values of the hypothesis tests whether the predictors are associated with HbA1c, and for each predictor give a conclusion whether it is associated with HbA1c yes/no [3 points].

```{r}
# burdenS P-value is 0.29169, since p > 0.05, there is no statistically significant evidence that burdenS is associated with HbA1c in this dataset. So, we can say that burdenS has a very low estimate 0.00178, meaning it doesn't seem to has a statistically significant effect on blood sugar levels in this dataset.

# age2 P-value is < 2e-16, and most of the individual values starting from 4-5y to 16-17y show highly significant association. So, we can say that as children get older (from age 4 and up till 17), their HbA1c estimates increase compared to the youngest group (0-1y) which was the reference group for the age in this model. The peak of this increase is in 14-15y group.

# schichtz P-value is 0.00242 (Medium) and 1.25e-05 (High). So, both "medium" and "high" social status levels have p < 0.05. This shows a statistically significant association between social status and HbA1c. So, we can say that both medium and high social status are associated with a decrease in HbA1c because of the negative estimates -0.024 and -0.041 compared to the low social status reference group in this model.
```

c) Now create a binary variable for high/low HbA1c. Select a clinically meaningful cutoff. Then assign value 1 if the child has higher HbA1c compared to the cutoff, and 0 if lower HbA1c compared to the cutoff. [3 points] 

```{r}
# As 5.7% and 6.4% is considered prediabetic, I would take the cutoff = 5.7% 
# as the standard for at-risk patients, because they are children
# Creating the binary variable, I am using as.numeric because (HbA1c >= cutoff) 
# returns TRUE/FALSE, TRUE becomes 1, FALSE becomes 0
kiggs_Ab_imp_com$HbA1c_bin <- ifelse(kiggs_Ab_imp_com$HbA1c >= 5.7, 1, 0)

table(kiggs_Ab_imp_com$HbA1c_bin)
# This shows high imbalance between the two groups which from my opinion is normal for random groups of children.
```

d) Calculate one logistic regression model with all variables as predictors and test in this model which of these 3 variables have an association with the overall HbA1c. [3 points] 

```{r}
# Running the logistic regression model with all variables as predictors
logit_fit <- glm(HbA1c_bin ~ as.numeric(age2) + schichtz + burdenS + as.numeric(sex) +
                   as.numeric(e065z), data = kiggs_Ab_imp_com, 
                 family = binomial(link = "logit"))

# View the results
summary(logit_fit)

# None of the chosen 3 variables have an association with the overall HbA1c. It seems that with only 406 as "1", the signal is too weak to be captured in this logisitic regression model.

```

e)
1- Report the estimated odds ratios for each predictor in the model, as well as the p-values from significance tests [3 points]. 

```{r}
jtools::summ(logit_fit, exp = T, confint = T, model.fit = F, digits = 3)

# According to results of using all variables, "Sex" shows the only statistically significant predictor in the binary model (p = 0.0225) but it isn't one of the 3 chosen variables.

#                                   Odds Ratio | P-values
#---------------------------------- ----------  --------
#(Intercept)                            0.052   0.000
#as.numeric(age2)                       1.013   0.524
#schichtzmedium social status           1.044   0.719
#schichtzhigh social status             0.916   0.551
#burdenS                                0.966   0.177
#as.numeric(sex)                        0.792   0.021
#as.numeric(e065z)                      0.979   0.370
#-----------------------------------------------------
```

2- Also give estimates of the 95% confidence interval of the odds ratios for each predictor in the model [3 points].  

```{r}
# Odds Ratios calculation
OR <- exp(coef(logit_fit))

# 95% Confidence Intervals calculation
CI <- exp(confint.default(logit_fit))

# Combining into a clean table
round(cbind(Odds_R = OR, Lower_CI = CI[,1], Upper_CI = CI[,2]),3)
```

3- Interpret the results (statistically significant relationship yes/no) for each predictor in the model and also interpret the odds ratios. [4.5 points]. 

```{r}
# let's start with the chosen variables first: 
# age2 isn't statistically significant as p = 0.524.

# OR Interpretation: The OR is 1.013. This means that for every 1 unit increase in the age category (as example from 4-5y to 6-7y), the odds of high HbA1c increase by 1.3% (very small change), holding all other variables constant.

#---------------------------------------------------
# schichtz (Social Status) isn't statistically significant as medium (p = 0.719) and high (p = 0.551).

# OR Interpretation: Those with Medium status have higher 4.4% odds (OR = 1.044) compared to the reference group (Low Social Status), and those with High social status have 8.4% lower odds (OR = 0.916) of high HbA1c.

#---------------------------------------------------
# burdenS (p = 0.177). Since p > 0.05, this relationship is not statistically significant.

# OR Interpretation: The OR is 0.966. For every 1 unit increase on the burden scale, the odds of high HbA1c decrease by about 3.4%. This means that the relationship is negative.

#---------------------------------------------------
# The only statistically significant predictor in this model was (sex) which p = 0.021 less than 0.05. 

# OR interpretation: The sex variable OR is 0.792. This means that for a 1 unit increase in the sex variable (moving from female to male), the odds of having high HbA1c decrease by approximately 20.8% as 1 - 0.792 = 0.208, holding all other variables constant.

#---------------------------------------------------
# e065z with (p = 0.370). So it isn't statistically significant.

# OR Interpretation: The OR is 0.979, indicating a 2.1% decrease in the odds of high HbA1c per unit increase in e065z. This is not a significant association.

```

f) What do you think about these two types of analyses, and is there any that seems more reliable to you (or both, or none)? [1.5 points] 

```{r}
# I think that linear regression was more reliable in this specific context because it shows more statistical significance with the predictors. It detected patterns that could be used to predict the HbA1c values. 

# On the opposite, the logistic regression suffered from imbalanced inputs, and this led to less number of statistically significant predictors. The logistic Regression here might lead me to incorrect conclusions about social status and age groups and their relationship to blood sugar, because the model from - my point of view - wasn't sensitive enough to see the trend. Finally, it seems like the logistic regression needs more predictors to be reliable, or may be different model. 
```

Notes: (1) You might want to check the formatting of all variables. (2) If you have problems solving question c), you can request the solution so that you can work on the next questions.


5) Question 5 – Sample size calculation [10 points]
Here, the aim is to perform a sample size calculation for a new study. The aim of the study is to investigate the effect of going for a run 3x per week (yes/no) on HbA1c levels. HbA1c might be measured as a continuous or binary variable, see previous question, and you can choose.

a) State whether you use HbA1c as continuous or binary variable 

```{r}
# I am using HbA1c as continuous variable.
```

b) State a reasonable effect size measure. Look at the literature or think for yourself based on expert knowledge what effect size you would expect. State the effect size that you are assuming and explain why. [3 points]

```{r}
# The reasonable effect size measure here is the mean difference in HbA1c (%) between the running group and the control group, expressed as an absolute mean difference (e.g. −0.5 percentage points HbA1c) or as a standardized mean difference (Cohen’s d) based on that change.

# Since I have chosen HbA1c as a continuous variable, the natural effect size is the expected difference between group in mean HbA1c, measured in HbA1c percentage points (e.g. 7.5% vs 7.0%). I would have thought about ANOVA if I have had more than 2 groups.

# For sample size formulas that use standardized effects, this can be converted to a Cohen’s d by dividing the expected mean difference by the assumed common standard deviation of HbA1c. Example: if I expect a −0.5% absolute reduction in HbA1c with a standard deviation of 1.0%, this corresponds to d = 0.5, which means a moderate standardized effect.

# Randomized trials and meta‑analyses of aerobic or structured exercise in adults with type 2 diabetes generally show absolute HbA1c reductions around 0.4–0.9 percentage points compared with control, depending on weekly exercise volume and supervision.

# Running 3 times per week corresponds to a moderate aerobic dose, so for the sample size calculation I assume an absolute mean difference of −0.5 percentage points HbA1c between the running and control groups. Assuming an HbA1c standard deviation of 1.0%, this corresponds to a Cohen’s d of 0.5, as a moderate standardized effect, which is both clinically relevant and consistent with the mid‑range of effects reported in the literature: doi: 10.1001/jama.2010.1710

```

c) Choose an appropriate statistical model/test for the sample size calculation, which should relate to your chosen effect size, and explain why. [3 points] 

```{r}
# I choose a two-sample comparison of means between running and control groups, this means a two-sample t‑test (or, more precisely, a linear model for HbA1c with group as main predictor).

# The reason behind my choice is the effect size which is defined as the between‑group mean difference in HbA1c (continuous outcome). The natural corresponding test for a parallel two‑arm trial is a two-sample t‑test for independent means. 

```

d) Now compute the minimum necessary sample size for a power of 80% and a significance threshold of alpha = 0.05, for example by using a function in the R package pwr. What is the sample size? [3 points] 

```{r}
pwr::pwr.t.test(d = 0.5, sig.level = 0.05, power = 0.80, type = "two.sample",
                alternative = "two.sided")

# The sample size n = 63.8 per group, rounded to 64 sample per group 
# The total sample size = 128 participants.

# To have an 80\% chance of detecting a medium effect size (d = 0.5) at a significance level of 0.05, I need to recruit a total of 128 participants (64 in the running group and 64 in the control group).

```

e) Do you think this is a good study, or do you see any major weaknesses in the study design or sample size calculation? [1 point]

```{r}
# It is hard to say it is a good study because there are a lot of unknown details, as people who run 3 times a week usually have other healthy habits, they might eat better, sleep deeper, smoke less, or have lower BMI mean. If I just compare "runners" vs. "non-Runners," I won't know if the lower HbA1c is affected from the running or other factors. 

# Also, asking people if they run 3 times per week (Yes/No) is unreliable. From personal experience, people tend to over report healthy behaviors because of social desirability. A "Yes" from someone who jogs for 10 minutes is very different from a "Yes" from a marathoner, the pace, the intensity, the habits after and before running are different but the study treats them as the same.

# Another problem is the ambiguity of the question itself, If some participants ran 2 times per week, will putting them in "NO" group is a suitable way? Even running 4 times or more per week could be answered "NO", which could lead to misleading results.

# Furthermore, medically, HbA1c measures average blood sugar over the past 2–3 months. If this study doesn't ensure that the "Runners" have been running consistently for at least 12 weeks before the blood draw, the results won't be accurate.

# The calculation of the required sample size says I need 128 people. But, in lifestyle studies, people drop out, stop running, or forget to show up for their blood tests. If I recruit exactly 128 and 10 people leave (which is highly expected even within highly motivated group), the power drops below 80%, and the study won't achieve the aim. I would recommend recruiting +150 participants if the other issues are solved.

# Finally, I am thinking about the effect of age and genetic variances of the 128 participants. The age could be controlled in the beginning of the study but strict population slice, but the genetic variance will stay. I don't know if assuming same genetics is an option or not but I think They aren't that big group that could dissolve the differences. So, the desired effect could appear or not, it depends on many other factors, but may be I went too far in this last point.
```

6) Question 6 – Meta analysis [20 points] 

Here, the aim is to perform a meta analysis.

a) Let’s continue to look at HbA1c as an outcome, still as you choose as continuous or binary variable. State your choice.

```{r}
# I will continue to treat HbA1c as a continuous variable.
```

b) Choose an exposure variable that you are interested in. It can be a continuous, 
categorical, ordinal, or binary variable. Which one have you chosen? [1 point] 

```{r}
# I choose body mass index (BMI) as a continuous exposure. This allows estimation of how much HbA1c changes per 1 unit or 5 units increase in BMI and avoids loss of information from inaccurate categorization.
```

c) Choose a population of interest. [1 point] 

```{r}
# I choose adults from both biological genders with diabetes.
```

d) Choose which study design (RCT, prospective cohort study, any study design …) the studies should have that investigated this association. [1 point] 

```{r}
# I choose cohort studies that follow adults with diabetes over time and assess the association between BMI (measured repeatedly) and HbA1c as continuous outcome. 
```

e) Search and choose 10 studies that have investigated the association according to your criteria. Screen the studies to make sure that they are relevant, and name them in your answer (eg title, author, year, journal, doi). [3 points] 

Answer:
I started my study with "prospective cohort studies on adult population with type 2 diabetes in Germany" but I couldn't find enough studies that match these specifications so, I widened my search to include all "cohort studies", all "diabetic" and wherever the patients are in the world. Starting with pubmed using this query: "(BMI OR "Body Mass Index") AND (HbA1c OR "Glycated Hemoglobin") AND (diabetes OR "Diabetes Mellitus") AND ("prospective cohort" OR "longitudinal study") AND (regression OR "beta coefficient" OR "standard error")",  with Filters applied: Abstract, Free full text, Full text, Observational Study, English, Humans, Adult: 19-44 years, Adult: 19+ years, Middle Aged: 45-64 years, Middle Aged + Aged: 45+ years, Young Adult: 19-24 years, MEDLINE. The studies I found weren't relevant to my search.

So, I used AI to help me gathering pool of paper after many trials to use pubmed and google scholar. The problem I faced is lack of parameters and the inaccuracy of results. It isn't very good with AI too but I think it has made it easier for trying different study designs to find a suitable one. I tried https://scispace.com/  free version using this prompt: 

"Use PubMed, Scispace Deep Search, and clinicaltrials.gov to find papers and clinical trials related to investigating the association between BMI (as a continuous exposure) and HbA1c (as continuous outcome) in "prospective cohort" specifically. make this investigation on "adults" from both biological genders with "diabetes"." I edited this prompt later and removed "prospective" with no much improvement.

This specific tool automated the designing of pubmed query with each trial and a suitable query for each database, but the results also lacked accuracy and required deep checking.

```{r}
# 1. Title: Distinct trajectories of HbA1c in newly diagnosed Type 2 diabetes from the DPV registry using a longitudinal group-based modelling approach
# Authors: W. Rathmann, A. Schwandt, J. Hermann, O. Kuss, M. Roden, K. Laubner, F. Best, S. Ebner, M. Plaumann, and R. W. Holl 
# Year: 2019 
# Journal: Diabetic Medicine 
# DOI: 10.1111/dme.14103 

#---------------------------------------------
# 2. Title: HbA1c trajectories over 3 years in people with type 2 diabetes starting second-line glucose-lowering therapy: The prospective global DISCOVER study 
# Authors: Brenda Bongaerts, Oliver Kuss, Fabrice Bonnet, Hungta Chen, Andrew Cooper, Peter Fenici, Marilia B. Gomes, Niklas Hammar, Linong Ji, Jesús Medina, Antonio Nicolucci, Kamlesh Khunti, Marina V. Shestakova, Hirotaka Watada, and Wolfgang Rathmann
# Year: 2023 
# Journal: Diabetes, Obesity and Metabolism 
# DOI: 10.1111/dom.15050

#---------------------------------------------
# 3. Title: Rates of glycaemic deterioration in a real-world population with type 2 diabetes 
# Authors: Louise A. Donnelly, Kaixin Zhou, Alex S. F. Doney, Chris Jennison, Paul W. Franks, Ewan R. Pearson 
# Year: 2018 (Published online 2017) 
# Journal: Diabetologia 
# DOI: 10.1007/s00125-017-4519-5

#---------------------------------------------
# 4. Title: Factors affecting Hemoglobin A1c in the longitudinal study of the Iranian population using mixed quantile regression 
# Authors: Abbas Bahrampour & Saiedeh Haji-Maghsoudi 
# Year: 2023 
# Journal: Scientific Reports 
# DOI: 10.1038/s41598-023-36481-x 

#---------------------------------------------
# 5. Sociodemographic factors associated with HbA1c variability in type 2 diabetes: a prospective exploratory cohort study
# Authors: Emelia Mellergård, Per Johnsson, Frida Eek 
# Year: 2020 
# Journal: BMC Endocrine Disorders 
# DOI: 10.1186/s12902-020-00585-6 

#---------------------------------------------
# 6. Title: Determinants of glycaemic control in a practice setting: the role of weight loss and treatment adherence (The DELTA Study) 
# Authors: C. McAdam-Marx, B. K. Bellows, S. Unni, J. Mukherjee, G. Wygant, U. Iloeje, J. N. Liberman, X. Ye, F. J. Bloom, D. I. Brixner 
# Year: 2014 
# Journal: International Journal of Clinical Practice 
# DOI: 10.1111/ijcp.12459

#---------------------------------------------
# 7. Title: Changes in body mass index, glycosylated hemoglobin A1C, blood pressure, and LDL-cholesterol among type 2 diabetes patients in Malaysia: A population-based longitudinal study
# Authors: Kim Sui Wan, Mastura Ismail, Foong Ming Moy, Feisul Idzwan Mustapha, Noran Naqiah Hairi 
# Year: 2021 
# Journal: Journal of Diabetes 
# DOI: 10.1111/1753-0407.13206

#---------------------------------------------
# 8. Title: Differential association of body mass index on glycemic control in type 1 diabetes
# Authors: Eun Young Lee, Yong-ho Lee, Sang-Man Jin, Hae Kyung Yang, Chang Hee Jung, Cheol-Young Park, Jae Hyoung Cho, Woo Je Lee, Byung-Wan Lee, Jae Hyeon Kim
# Year: 2016
# Journal: Diabetes/Metabolism Research and Reviews
# DOI: 10.1002/dmrr.2815

#---------------------------------------------
# 9. Title: Associations between decreases in adiposity and reductions in HbA1c and insulin use in the Look AHEAD cohort 
# Authors: Rebecca Palmer, Rebecca H.Neiberg, Kristen M.Beavers, Steven E. Kahn, Denise K.Houston, Lynne Wagenknecht, Karen C.Johnson, Henry J.Pownall, Mark A. Espeland
# Year: 2025
# Journal: Obesity (Silver Spring)
# DOI: 10.1002/oby.24242

#---------------------------------------------
# 10. Title: Assessing related factors to fasting blood sugar and glycosylated hemoglobin in patients with type 2 diabetes simultaneously by a multivariate longitudinal marginal model
# Authors: Samaneh Hosseinzadeh, Zahra Khatirnamani, Enayatollah Bakhshi, Alireza Heidari, Arash Naghipour 
# Year: 2022 
# Journal: Scientific Reports
# DOI: 10.1038/s41598-022-19241-1 
```

f) Now perform a meta-analysis for computing a pooled effect. Report all the steps in the preparation (create a data frame with the study info and results, transform the results if necessary, choose a fixed or random effect model), the pooled effect and its 95% confidence interval, and compute and show a nicely formatted forest plot of the results. [12 points]

Answer:

Some studies has OR instead of the beta coeffecient so, I have no other solution except a try of conversion first to complete the required 10 studies. Others categorized BMI or HbA1c, so some calculations are needed before insertion in the metafor code. So, evaluation of studies quality and relevancy were mentioned before each study below using AI "Gemini":

The 1st study Overall Quality: Large sample size (n=6355), longitudinal design, and use of a high-quality multicentre registry (DPV). HbA1c values were standardized to reduce laboratory variation. But its relevancy to the meta analysis isn't high as it uses Group-Based Trajectory Modeling (GBTM). This technique identifies "clusters" of people. Instead of saying "HbA1c goes up by X," they say "This person belongs to the 'Stable, Poor, Control' group.

round(log(1.05), 4)
round((log(1.07) - log(1.03)) / 3.92, 4)

---------------------------------------------
The 2nd study Overall Quality: It features a global sample size (n=9295) across 37 countries, providing high external validity. Also, its relevancy to the meta analysis isn't high as it uses Latent Class Growth Modelling (LCGM). It groups people into "trajectories." and uses Multinomial Logistic Regression and reported OR.
round(log(1.04), 4)
round((log(1.10) - log(0.99)) / 3.92, 4)

---------------------------------------------
The 3rd study Overall Quality: Large sample size (n > 5,000), long follow-up duration (median 9.4 years), and robust adjustment for time-dependent confounders including medication changes and glucocorticoid use. The study specifically highlights that only the highest BMI category showed a statistically significant independent association with a higher rate of glycaemic deterioration in the multivariate model.
round((0.47 - 0.06) / 3.92, 4)

---------------------------------------------
The 4th study Overall Quality: The study features a large, well-defined cohort (N=2791) with a long follow-up duration (up to 13 years), providing good statistical power. The use of mixed-effects modeling appropriately accounts for the longitudinal nature of the data (repeated measures). It is included here with Caution. While the study meets all primary inclusion criteria, it utilizes Linear Mixed Quantile Regression (LQMM) rather than a standard mean-based regression. For a standard meta-analysis, the beta at the 50th quantile (median) is typically used as the best representative of the central tendency when mean effects are not reported.

---------------------------------------------
The 5th study Overall Quality: this study does not provide a beta coefficient for BMI as a continuous variable. The researchers treated BMI as a categorical variable (Normal, Overweight, Obese) in their Analysis of Variance (MANOVA). They used ANOVA-based models rather than linear regression with BMI as a continuous scale. They report the Mean Difference between specific categories.

I use the Mean Difference between the Obese and Normal groups as the effect size (yi): 0.34
# And To find the SE from the 95% CI (−0.04, 0.71)
(0.71 - (-0.04)) / 3.92 = 0.75/3.92= 0.191 #SE

---------------------------------------------
The 6th study Overall Quality: It uses a low sample size (N=477) compared to others and employs Structural Equation Modeling (SEM), which allows for the simultaneous assessment of complex, interrelated variables like weight loss, medication adherence, and glycemic control. It introduce OR instead of beta coefficient.
the beta (slope) for the association between baseline BMI and HbA1c level is 0.
(log(1.03) - log(0.97)) / 3.92 # SE

---------------------------------------------
The 7th study Overall Quality: The study is of high quality for a population-based longitudinal analysis. With over 18,000 patients, it has high statistical power. It utilizes the National Diabetes Registry, covering all public health clinics in a Malaysian state, which is comparable to national averages. It tracks individual changes by merging annual clinical audit datasets. While the study reports the mean changes for both BMI and HbA1c over the follow-up period, it does not explicitly report a regression coefficient or a direct correlation specifically for the association between the change in BMI and the change in HbA1c in a way that provides a single effect size for that relationship. So to calculate the required values for the meta-analysis, I extracted the mean and standard deviation (SD) for the total study population (n = 18341) at baseline and study exit from the longitudinal audit data presented in Table 2.

yi: 8.04 - 7.88 = 0.16
SEi: sqrt((2.03^2 / 18315) + (2.15^2 / 18315))= 0.0219

---------------------------------------------
The 8th study Overall Quality: Small sample size (N=719) for a Type 1 Diabetes study; with multi-centre design; using 18-month average HbA1c (which reduces biological noise compared to a single measurement). The study relied only on baseline BMI. Without longitudinal BMI data, it cannot confirm if changes in weight caused the changes in HbA1c.
I calculated the Pooled beta (yi) Then, calculate the pooled standard error as the square root of the reciprocal of the sum of the weights (Using AI assistance).

---------------------------------------------
The 9th study Overall Quality: It utilizes data from a well-characterized, large-scale clinical trial cohort (Look AHEAD) with long-term (8-year) follow-up. The authors used generalized linear mixed models adjusted for baseline levels and repeated measurements, providing a stable estimate of the association over time.

Effect size (beta): -0.13 (representing the mean change in HbA1c 1% for every 5% reduction in BMI) and to make it to 1% of BMI:
(-0.13/5 = -0.026)
(0.16 - 0.11) / (2 * 1.96) = 0.0128/5 = 0.0025 # Calculated Standard Error (SE) Derived from the 95% CI

---------------------------------------------
The 10th study Overall Quality: Although, the meta-analysis specifically requires BMI as a continuous variable, this study records and analyzes "Weight at first visit (kg)". While the authors mention in their discussion that no significant relationship was observed between "body weight (body mass index)" and HbA1c, the actual data extracted and modeled was body weight in kilograms, not the BMI ratio (kg/m^2).


```{r}
library(metafor)

# Create data frame based on selected studies
meta_data <- data.frame(
  study = c("Rathmann et al.", "Bongaerts et al.", "Donnelly et al.", "Bahrampour et al.", "Mellergård et al.", "McAdam-Marx et al.", "Wan et al.", "Lee et al.", "Palmer et al.", "Hosseinzadeh et al."),
  year = c(2019, 2023, 2018, 2023, 2020, 2014, 2021, 2016, 2025, 2022),
  # Effect size (Beta)
  yi = c(0.0488, 0.0392, 0.26, 0.0233, 0.34, 0, 0.16, 0.052, -0.026, 0.13), 
  # Standard Error
  sei = c(0.0097, 0.0269, 0.1046, 0.0062, 0.191, 0.0153, 0.0219, 0.037, 0.0025, 0.08) 
)

# Sampling variance (vi) calculation
meta_data$vi <- meta_data$sei^2

# Fitting random-effects model
res <- rma(yi, vi, data = meta_data, method = "REML")
summary(res)

# Forest plot:
forest(res, slab = paste(meta_data$study, meta_data$year, sep = ", "), 
       xlab = "Change in HbA1c per unit BMI", 
       main = "Association between BMI and HbA1c")

```
g) Name 2 weaknesses of this meta-analysis. Do you think this could still be representative of a real meta-analysis on the topic? [2 point]

```{r}
# 1. The major problem I am facing is mixing mean differences from categorical studies with beta coefficients from continuous variables studies in the same meta-analysis. This has introduced the possibility of higher bias. For better statistical approach, I think only using studies that report the linear regression slope (beta) to ensure the units (change per 1 kg/m^2) are identical across the forest plot, but this wasn't available in 10 studies.

# 2. A problem I noticed in longitudinal studies is the difficulty of maintaining the followup period or the units of the variables across the selected studies. In some studies, it is 18 months, and in other studies, it is more than 3 years. In biological lifestyle variables, the continuity could improve the results with long term implications. 

# Generally, I won't recommend taking the approach I used except in extreme cases, like exams XD
```

Hint: in the creation of the necessary dataset for the analysis, you can take the datasets provided in the metafor package as reference how the dataset should be formatted. If you have problems with dataset formatting, you could still solve the question by hand, i.e. compute the pooled effect by hand, and draw the forest plot by hand (and insert a picture in your output report file) to get some of the available points.